---
title: "Non And Semi Parametric Survival Analysis - DrPH(Epid)"
author: "Kamarul Imran Musa"
date: "18 April 2016"
output: 
  pdf_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Cox proportional hazard model

## Introduction

### The coxph function

The Cox proportional-hazards regression model can be fit in R with the `survival::coxph` function.

The argumen *method* indicates how to handle observations that have tied (i.e., identical) survival times. The default "efron" method is generally preferred to the once-popular "breslow" method.

# Initial preparation

## Prepare folder and file

Set the working directory to store all the working datasets and also the outputs. 

```{r}
setwd("E:/Epi_Stat_Matters/LectureNotes2015/Survival-Analysis/survival-analysis-DrPH-epid-2015/Practicals/Survival_Analysis_DrPH/NonAndSemiParametricSurvivalAnalysis")

list.files()
```

## Read data

We can use `foreigh::read.dta` to read stata format data. 

```{r readdata}
library(foreign)
data1<-read.dta('addicts.dta',convert.factors = T)
```

## Load library 

The popular package to run survival analysis in R is `survival`package. We can load the package using the `library` function

```{r}
library(survival)
```

# Preliminary analysis

## Declare file as a survival data format

It is important to tell R that the data for survival analysis is in the time-to-event format. We need to specify:

1.  variable time
2.  variable event. What would be the event of interest. The rest will be censored observations. 

When can see the summary of the data

```{r}
datas <- Surv(time = data1$survt,event = data1$status == 1)
summary(datas)

```

## Overview of data

Let us see the first 50 observations for the 

1.  time-to-event data
2.  the original data

```{r}
head(datas,50)
head(data1,50)
```

+ sign is for censored observation

# Estimation of the survival based on non-parametric method

To estimate the expected survival probability, we can use the non-parametric method. The most popular one is the Kaplan-Meier survival estimate. 

## Estimate the crude (unadjusted) survival functions for all 

Using the intercept-only model to obtain Kaplan-Meier survival estimates for all event times. 

```{r kaplan1}
surv.fit<-survfit(datas~1, data = data1)
summary(surv.fit)
```

And also for a specific time for example the expected survival at time 100.

```{r kaplan2}
summary(surv.fit,times = 100)
```

## Estimate crude (unadjusted) survival functions for strata

For each clinic

```{r kaplan3}
surv.clin<-survfit(datas~clinic, data = data1)
summary(surv.clin)
```

For time = 100

```{r kaplan4}
summary(surv.clin,times = 100)
```

## Kaplan-Meier survival plots 

We will plot the survival probability against time for all observations

```{r}
plot(surv.fit, xlab='survival times (days)', ylab='survival prob')
```

We will plot the survival probability against time for observations based on clinic.

```{r}
str(data1$clinic)
str(data1$prison)
```

Let us plot. Note that the *solid* line and *black* color belong to clinic1 and the *dashed* line with *red* color belong to clinic2 

```{r plotnolegend}
plot(surv.clin, lty=c('solid','dashed'), col=c('black','red'))
```

Now, we add legend

```{r addlegend}
plot(surv.clin, lty=c('solid','dashed'), col=c('black','red'))
legend('topright',c('clinic 1','clinic 2'),lty=c('solid','dashed'), col=c('black','red'))
```

## Inferences based on Kaplan-Meier survival estimates

### The log rank test

To test for the difference in the estimated survival times by the Kaplan-Meier, the default is the log-rank test

```{r}
survdiff(datas~clinic, data = data1)
```

# Estimation using the Semi-parametric method

The most common semi-parametric method to estimate the survival probability is the Cox proportional hazard regression model

## The Cox PH model

Efron is default method in R to estimate the survival based on the Cox PH model

```{r}
data1.cox <- coxph(datas ~ prison + dose + clinic, data = data1)
summary(data1.cox)

```

column **z** is the ratio between each regression coefficient to its SE. It is a Wald statistic which is asymptotically standard normal under the hypothesis that the corresponding beta is 0.

Other alternatives = Breslow , Exact

Exponentiated coefficients in the second column of the first panel (and in the first column
of the second panel) of the output are interpretable as multiplicative effects on the hazard or simply as **Hazard Ratio**

## Estimated distribution of survival times based on Cox PH model

The survfit function estimates S(t), by default at the **mean values of the covariates**.

```{r}
survfit(data1.cox)
summary(survfit(data1.cox))
```

The plot method for objects returned by survfit graphs the estimated surivival function, along
with a point-wise 95-percent confidence band.

```{r}
plot(survfit(data1.cox),
     xlab = 'times', ylab = 'proportion status == 1')
```



## Checking the PH assumption

Tests and graphical diagnostics for proportional hazards may be based on the scaled Schoenfeld
residuals; these can be obtained directly as *residuals(model, "scaledsch")*, where model is a
*coxph* model object. The matrix returned by residuals has one column for each covariate in
the model. More conveniently, the cox.zph function calculates tests of the proportional-hazards
assumption for each covariate, by correlating the corresponding set of scaled Schoenfeld residuals

### Using graphical methods

We will use *cox.zph*

It computes a test for each covariate, along with a global test for the model as a whole

```{r}
cox.zph(data1.cox)
```

Strong evidence of non-proportional hazards for the whole model and for covariate *age*

Get the detailed plots after *cox.zph*. Plotting the object returned by cox.zph produces graphs of the scaled Schoenfeld residuals against transformed time.  Systematic departures from a horizontal line are indicative of
non-proportional hazards.

```{r}
plot(cox.zph(data1.cox))

```



## Modification if a model violates PH assumption 

An alternative to incorporating an interaction in the model is to divide the data into strata based
on the value of one or more covariates. Each stratum is permitted to have a different baseline hazard
function, while the coefficients of the remaining covariates are assumed to be constant across strata.
An advantage of this approach is that we do not have to assume a particular form of interaction
between the stratifying covariates and time. A disadvantage is the resulting inability to examine
the effects of the stratifying covariates. Stratification is most natural when a covariate takes on
only a few distinct values, and when the effect of the stratifying variable is not of direct interest

Run, the strata argument for clinic (which has threaten the proportionality)

```{r}
str.cox <- coxph(datas ~ prison + dose + strata(clinic),
                 data = data1)
summary(str.cox)
```

Any existing violation of PH assumption

```{r}
cox.zph(str.cox)
```

Nope. All are good. All fulfill PH assumption. 

## Cox PH model with interaction term

Read page 197 and 198 Applied Survival Analysis by Hosmer and Lemeshow. 

# Prediction

## Predict relative risk 

We use `predict.coxph` to predict the outcomes after running the Cox PH model. For example to obtain the relative risk; that is the risk of population of interest (with a set of covariates) against the population average, we can use these:

```{r predictrisk}
pred_risk <- predict(data1.cox, type = 'risk')
cbind(head(data1[, c('prison', 'dose', 'clinic')]), 
      head(pred_risk))

```

## Plot the expected survival probability after Cox model

We can plot the survival probability against time based on the Cox PH model 

```{r plotcox}
plot(survfit(data1.cox), col=4, 
     xlab = 'time', ylab = 'survival prob', main = 'survival prob against time based on Cox model')

```


# Model assessment

## Influential observation

Check the dfbeta values

```{r}
dfbeta <- residuals(data1.cox, type="dfbeta")
par(mfrow=c(2, 2))
for (j in 1:3) {plot(dfbeta[, j], ylab=names(coef(data1.cox))[j])
abline(h=0, lty=2)
}

```

Summarize the dfbetas values

```{r}
names(coef(data1.cox))
summary(dfbeta)
```

Plot the dfbetas values

```{r}
plot(dfbeta[,1]) #prison
plot(dfbeta[,2]) #dose
plot(dfbeta[,3]) #clinic

```

# Additional tutorial

method 1

```{r}
plot(surv.clin,fun='cloglog',
    xlab='time(days) using log scale', ylab='log-log survival prob',
    main='log-log curves by clinics')
```


Looks the curves cross each other. This indicate model violate PH assumption

## Using statistical test

```{r}
test.ph <- coxph(datas ~ prison + dose + clinic, data = data1)
test.ph2 <- cox.zph(test.ph, transform = rank)
test.ph2
# var = clinic means residuals should pertain to the variable clinic
plot(test.ph2, se=FALSE, vsr = "clinic")

```

# Running stratified Cox model

When models violate PH assumption. in our case, 'clinic' does violate PH assumption but others are not. So we do stratified Cox model

```{r}
surv.strata <- coxph(datas ~ prison + dose + strata(clinic), data = data1)
summary(surv.strata)
```

# Running stratified Cox model with interaction

```{r}
surv.strata.ia <- coxph(datas ~ prison + dose + clinic:dose + clinic:prison + strata(clinic), data=data1)
summary(surv.strata.ia)
```

Calculating the HR

If we want to calculate the HR betwwen PRISON=1 vs PRISON=0 for CLINIC=2, then one way is by making CLINIC equals 0. So when CLINIC==2, then CLINIC2==0. 

```{r}
data1$clinic2<- (as.numeric(data1$clinic)) - 1
summary(data1$clinic2)
head(as.numeric(data1$clinic))
surv.strata.ia2 <- coxph(datas ~ prison + dose +  clinic2:dose + clinic2:(as.numeric(prison)) +
                       strata(clinic2),data=data1)
summary(surv.strata.ia2)
```

# References:

1.  <https://socserv.socsci.mcmaster.ca/jfox/Books/Companion/appendix/Appendix-Cox-Regression.pdf>
2. Reference for calculating the relative risk <http://stats.stackexchange.com/questions/44896/how-to-interpret-the-output-of-predict-coxph>
3.  Another reference for predicting the outcomes after Cox model <http://datamining.togaware.com/survivor/Lung1.html>
