---
title: "Non And Semi Parametric Survival Analysis"
author: "Kamarul Imran Musa"
date: "18 April 2016"
output: 
  pdf_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare folder and file

```{r}
setwd("E:/Epi_Stat_Matters/LectureNotes2015/Survival-Analysis/survival-analysis-DrPH-epid-2015/Practicals/NonAndSemiParametricSurvivalAnalysis")
list.files()
```

# Read data

```{r}
library(foreign)
data1<-read.dta('addicts.dta',convert.factors = T)
```

# Load library survival

```{r}
library(survival)
```

# Declare file as a survival data format

```{r}
datas<-Surv(time=data1$survt,event=data1$status==1)
summary(datas)
head(datas,50)
head(data1,50)
```

+ sign is for censored observation

# Estimate crude (unadjusted) survival functions for all 

Using intercept-only model to obtain Kaplan-Meier survival estimates for all event times. And also for a specific time.

```{r}
surv.fit<-survfit(datas~1)
summary(surv.fit)
summary(surv.fit,times = 100)
```

# Estimate crude (unadjusted) survival functions for strata

```{r}
surv.clin<-survfit(datas~clinic, data = data1)
summary(surv.clin)
summary(surv.clin,times = 100)
```

# Plots

```{r}
plot(surv.fit, xlab='survival times (days)', ylab='survival prob')
plot(surv.clin, lty=c('solid','dashed'), col=c('black','red'))
  legend('topright',c('clinic 1','clinic 2'),lty=c('solid','dashed'), col=c('black','red'))
```

# Check for survival difference

Default is log-rank test

```{r}
survdiff(datas~clinic, data = data1)
```

# Cox PH model

Efron is default method

```{r}
data1.cox<-coxph(datas~prison+dose+clinic,data=data1)
summary(data1.cox)
```

Other alternatives = Breslow , Exact

# Checking PH assumption

## Using graphical methods

Check PH assumption for clinic

method 1

```{r}
plot(surv.clin,fun='cloglog',
    xlab='time(days) using log scale', ylab='log-log survival prob',
    main='log-log curves by clinics')
```

method 2

```{r}
surv.clin2<-summary(surv.clin)
surv.clin2
surv.clin3<-data.frame(surv.clin2$strata,surv.clin2$time,surv.clin2$surv)
surv.clin3
#give column names
colnames(surv.clin3)<-c('clinic','time','survival')
surv.clin3[1:5,]

#stratify
clinic1<-surv.clin3[surv.clin3$clinic=='clinic=1',]
clinic1
clinic2<-surv.clin3[surv.clin3$clinic=='clinic=2',]
clinic2

plot(clinic1$time,log(-log(clinic1$survival)),xlab='survival time in days',ylab='log-log survival',col='blue', type='l',lty='solid')
#overlay plots
par(new=T)
plot(clinic2$time,log(-log(clinic2$survival)),xlab='survival time in days',ylab='log-log survival',col='red', type='l',lty='dashed')
#back to default
par(new=F)
legend('bottomright', c('Clinic1', 'Clinic2'), lty = c('solid', 'dashed'),col=c
('blue','red'))

```

Looks the curves cross each other. This indicate model violate PH assumption

## Using statistical test

```{r}
test.ph<-coxph(datas~prison+dose+clinic,data=data1)
test.ph2<-cox.zph(test.ph,transform = rank)
test.ph2
# var = clinic means residuals should pertain to the variable clinic
plot(test.ph2,se=FALSE,var='clinic')

```

# Running stratified Cox model

When models violate PH assumption. in our case, 'clinic' does violate PH assumption but others are not. So we do stratified Cox model

```{r}
surv.strata<-coxph(datas~prison+dose+strata(clinic),data=data1)
summary(surv.strata)
```

# Running stratified Cox model with interaction

```{r}
surv.strata.ia<-coxph(datas~prison+dose+clinic:dose+clinic:prison+strata(clinic),data=data1)
summary(surv.strata.ia)
```

Calculating the HR

If we want to calculate the HR betwwen PRISON=1 vs PRISON=0 for CLINIC=2, then one way is by making CLINIC equals 0. So when CLINIC==2, then CLINIC2==0. 

```{r}
data1$clinic2<-data1$clinic-1
head(data1$clinic2)
head(data1$clinic)
surv.strata.ia2<-coxph(datas~prison+dose+clinic2:prison+clinic2:dose+
                       strata(clinic2),data=data1)
summary(surv.strata.ia2)
```